Safe Memory Regions for Big Data Processing

### 2016-06-08 paper/type-inference.tex
type inference

@@

\begin{definition}
\emph{(\textbf{Constraint Satisfaction Problem (CSP)})}
A constraint satisfaction problem is a tuple
$(\bar{\pi}, \bar{\rho}, \bar{\varphi}, \regionDeltaMap, \predDeltaMap, C)$,
where $\bar{\pi}$ is a set of region constants,
$\bar{\rho}$ is a set of region variables,
$\bar{\varphi}$ is a set of predicate variables,
$regionDeltaMap : \bar{\rho} \rightarrow 2^{\bar{\pi}}$ is a function that
specifies a unification domain for each region variable,
$\predDeltaMap : \bar{\varphi} \rightarrow 2^{\bar{\pi}}$ is a function that
specifies the domain for each predicate variable,
and $C$ is a set of validity constraints, where the $i$'th validity constraint
assumes one of the following forms:
\begin{center}
\(
    \isvalid{\phictxt^{i} \conj \varphi}
            {\phicstr^{i}}\qquad
    \isvalid{\phictxt^{i} \conj \varphi}
            {F_i(\varphi)}
\)
\end{center}
The solution to the constraint satisfaction problem is a pair $(\regionSubstFn,\predSubstFn)$,
where $\regionSubstFn$ is a map from $\rhobar$ to $\bar{\pi}$
and $\predSubstFn$ is a map from $\bar{\varphi}$ to a region-constraint formula such that
\begin{itemize}
  \item $\regionSubstFn(\rho) \in \regionDeltaMap(\rho)$, for every $\rho \in \bar{\rho}$,

@@

Solving the constraints is an iterative process, consisting of the following
steps, and this process fails at any point if it determines that no solution
exists for the set of constraints:

@@

\item
Consider any constraint of the form $\isvalid{\varphi}{\pi_i \outlives \pi_j}$,
where $\pi_i$ and $\pi_j$ are both region constants.
If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$, then we simply add
$\pi_i \outlives \pi_j$ as an additional conjunct to $\predSubstFn(\varphi)$.
Otherwise, the constraint solver fails, since no valid solution is possible.

\item
  When any outlives constraint $\phi$ is added to $\predSubstFn(\varphi)$,
  then for every existing validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\varphi)}$,
  we produce a new validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\phi)}$.

@@

\paragraph{Properties of generated constraints}
The set of generated constraints have certain properties that allow us to
solve the constraints efficiently.
%
Since $\FB$ does not admit null values or uninitialized variables, every region
variable is forced to be unified with some concrete region, by the constraints.

>>>>

\paragraph{Constraint Generation.}

@@

\begin{definition}
\emph{(\textbf{Constraint Satisfaction Problem (CSP)})}
A constraint satisfaction problem is a tuple
$(\bar{\pi}, \bar{\rho}, \bar{\varphi}, \regionDeltaMap, \predDeltaMap, C)$,
where $\bar{\pi}$ is a set of region constants,
$\bar{\rho}$ is a set of region variables,
$\bar{\varphi}$ is a set of predicate variables,
$\regionDeltaMap : \bar{\rho} \rightarrow 2^{\bar{\pi}}$ is a function that
specifies a unification domain for each region variable,
$\predDeltaMap : \bar{\varphi} \rightarrow 2^{\bar{\pi}}$ is a function that
specifies the domain for each predicate variable,
and $C$ is a set of validity constraints, where the $i$'th validity constraint
assumes one of the following forms:
\begin{center}
\(
    \isvalid{\phictxt^{i} \conj \varphi}
            {\phicstr^{i}}\qquad
    \isvalid{\phictxt^{i} \conj \varphi}
            {F_i(\varphi)}
\)
\end{center}
The solution to the constraint satisfaction problem is a pair $(\regionSubstFn,\predSubstFn)$,
where $\regionSubstFn$ is a map from $\rhobar$ to $\bar{\pi}$
and $\predSubstFn$ is a map from $\bar{\varphi}$ to a region-constraint formula such that
\begin{itemize}
  \item $\regionSubstFn(\rho) \in \regionDeltaMap(\rho)$, for every $\rho \in \bar{\rho}$,

@@

\paragraph{Properties of generated constraints}
The set of generated constraints have two interesting properties that allow us to
solve the constraints efficiently.
%
The first property is that every region variable is forced to be unified with some
region constant by the constraints, except in one special case.
This follows because $\FB$ does not admit uninitialized variables.
(The only case where this does not hold is in the case of a recursive function
that calls itself in a non-terminating fashion: in this case, the function never
returns a value, and so the return-value can be typed as anything. In the sequel,
we assume that the return value of such a function is typed to be \C{unit}.)
% , and the type system uniquely determines the region parameters of every value
%
The second property is that the \emph{context region constraint} $\phictxt$
occurring on the antecedent of any validity constraint is a conjunction of
outlives-constraints of the form $\pi \outlives \pi_s$ where $\pi_s$
is a static region identifier and $\pi$ is a region constant (either a formal
region parameter, or static region identifier, or open transferable region identifier).
Furthermore, if $\phictxt$ includes any conjunct $\pi \outlives \pi_s$, then
it includes every conjunct $\pi_f \outlives \pi_s$ for every $\pi_f \in \predDeltaMap(\varphi)$.

\paragraph{Constraint Solver.}
Solving the constraints is an iterative process, consisting of the following
steps, and this process fails at any point if it determines that no solution
exists for the set of constraints:

@@

\item
\label{item:context}
Consider any constraint of the form $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$,
where $\pi_i$ and $\pi_j$ are both region constants.
If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$, then we simply add
$\pi_i \outlives \pi_j$ as an additional conjunct to $\predSubstFn(\varphi)$.
Otherwise, we check if $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
If this entailment does not hold, the constraint solver fails, since no valid solution is possible.

\item
  When any outlives constraint $\phi$ is added to $\predSubstFn(\varphi)$,
  then for every existing validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\varphi)}$,
  we produce a new validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\phi)}$. 
\end{enumerate}

Point~\ref{item:context} makes use of the special structure of the generated
constraints: namely that $\phictxt$  is a conjunction of outlives-constraints of the
form $\pi \outlives \pi_s$ where $\pi_s$ is a static region identifier and $\pi$ is a
region constant. As a result, we can show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.

This iterative process is a standard fixed point computation process.
It is possible to encode this computation using a set of Datalog rules, and compute the
solution using any Datalog engine.
Alternatively, note that the above computation, in the absence of context constraints
$\phictxt$ on the left-hand-side of the entailment, is a special instance of context-free
reachability in graphs. Algorithms for context-free reachability can be adapted to
incorporate the above treatment of context constraints $\phictxt$.

@@
